---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  cache = FALSE
)
```


# airqualitytr

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/airqualitytr)](https://CRAN.R-project.org/package=airqualitytr)
[![R-CMD-check](https://github.com/emraher/airqualitytr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/emraher/airqualitytr/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

**airqualitytr** is an R package that provides an interface to download and explore air quality data from the Turkish Ministry of Environment, Urbanization and Climate Change. It allows users to:

- Download time series air quality data for a specified station and parameters.
- Retrieve the lists of available cities.
- Retrieve station metadata.
- Retrieve available measurement parameters.

These functions enable researchers and analysts to quickly access and analyze environmental data in their R workflow.

**Note:** This package was initially created by Emrah Er, with subsequent improvements made using [Claude Code](https://claude.com/claude-code).

## Installation

You can install the development version from GitHub using the **pak** package:

```{r, eval=FALSE}
pak::pak("emraher/airqualitytr")
```


## Usage

Load the package and explore the available options:

### List Available Cities

Retrieve a tibble of cities available for filtering:


```{r}
library(airqualitytr)
library(dplyr)

(cities <- list_cities())
```


### List Available Stations

Get metadata for all available stations:

```{r}
stations <- list_stations()
head(stations)
```


### List Available Measurement Parameters

Retrieve a tibble of available measurement parameters (e.g., PM10, CO, NO2):

```{r}
(params <- list_parameters())
```


### Download Air Quality Data

Download time series air quality data for a specific station. You must provide the station's unique identifier, select one or more measurement parameters, specify a date range, and choose the data frequency (hourly or daily).

Here is an example:

```{r}
# Example inputs:
station_id <- "468478b7-ace5-4bd3-b89a-a9c1c2e53080"
parameters <- c("PM10", "CO", "NO2")
start_datetime <- "2025-03-16 00:00"
end_datetime <- "2025-04-15 00:00"
frequency <- "hourly"

data <- download_air_quality_data(
  station_id = station_id,
  parameters = parameters,
  start_datetime = start_datetime,
  end_datetime = end_datetime,
  frequency = frequency
)
head(data)
```


The resulting tibble contains:

- **time:** POSIXct timestamp of the measurement.
- **station_id:** Station identifier from the API response.
- **city_name:** The city name provided.
- **station_name:** The station name provided.
- **parameter:** The measurement parameter (e.g., "PM10").
- **value:** The corresponding measurement value.


### Bulk Download

```{r, eval=FALSE}
# Define the date range and parameters
start_datetime <- "2025-03-16 00:00"
end_datetime <- "2025-03-16 01:00" # using a short interval for testing
parameters <- c("PM10")
frequency <- "hourly"

# Bulk download air quality data for all stations
bulk_data <- bulk_download_air_quality_data(
  start_datetime = start_datetime,
  end_datetime = end_datetime,
  parameters = parameters,
  frequency = frequency
)

# Inspect the combined data
glimpse(bulk_data)
```

### Parallel Bulk Downloads

For faster bulk downloads, you can use parallel processing. This is especially useful when downloading data from many stations or large date ranges:

```{r, eval=FALSE}
# Install parallel processing packages if not already installed
# install.packages(c("future", "furrr", "progressr"))

library(future)
library(progressr)

# Set up parallel processing with 4 workers
plan(multisession, workers = 4)

# Bulk download with parallel processing
start_datetime <- "2025-03-16 00:00"
end_datetime <- "2025-03-16 01:00"
parameters <- c("PM10", "NO2")
frequency <- "hourly"

# Option 1: Let the function manage the parallel plan
bulk_data <- bulk_download_air_quality_data(
  start_datetime = start_datetime,
  end_datetime = end_datetime,
  parameters = parameters,
  frequency = frequency,
  parallel = TRUE,
  workers = 4 # Number of parallel workers
)

# Option 2: Set up your own future plan
plan(multisession, workers = 4)
bulk_data <- bulk_download_air_quality_data(
  start_datetime = start_datetime,
  end_datetime = end_datetime,
  parameters = parameters,
  frequency = frequency,
  parallel = TRUE # Uses your existing plan
)
plan(sequential) # Reset to sequential processing

# With progress tracking (requires progressr package)
handlers(global = TRUE) # Enable progress handlers
handlers("progress") # Use progress bar

with_progress({
  bulk_data <- bulk_download_air_quality_data(
    start_datetime = start_datetime,
    end_datetime = end_datetime,
    parameters = parameters,
    frequency = frequency,
    parallel = TRUE,
    workers = 4,
    show_progress = TRUE
  )
})

glimpse(bulk_data)
```

You can also use parallel processing for downloading all data from a single station:

```{r, eval=FALSE}
station_id <- "468478b7-ace5-4bd3-b89a-a9c1c2e53080"
parameters <- c("PM10", "CO", "NO2")

# Download all available data in parallel (parallelizes across time chunks)
plan(multisession, workers = 4)
all_data <- download_all_data_for_station(
  station_id = station_id,
  parameters = parameters,
  frequency = "hourly",
  chunk_size = "3 months", # Smaller chunks = more parallelization
  parallel = TRUE,
  workers = 4
)
plan(sequential)

glimpse(all_data)
```

### Bulk Download 2

```{r, eval=FALSE}
station_id <- "468478b7-ace5-4bd3-b89a-a9c1c2e53080"
parameters <- params$name
frequency <- "hourly"

# Download all available data for the specified station over the broad default range
all_data <- download_all_data_for_station(
  station_id = station_id,
  parameters = parameters,
  frequency = frequency
)

glimpse(all_data)
```

### Bulk Download 3 - Save All Station Data to Parquet Files

For very large bulk downloads, you can process each station sequentially or use parallel processing with `furrr`:

```{r, eval=FALSE}
library(tidyverse)
library(arrow)
library(future)
library(furrr)
library(airqualitytr)

# Define the parameters to download.
params <- list_parameters()
parameters <- params$name # or simply: c("PM10", "CO", "NO2")

# Retrieve station metadata from your package
stations_meta <- list_stations()

# Join city names (already included in list_stations() output)
cities_meta <- list_cities()
stations_meta <- stations_meta |>
  left_join(cities_meta, by = "city_id")

# Define frequency for downloading data.
frequency <- "hourly" # or "daily"

# Create an output folder to store Parquet files
if (!dir.exists("bulk_station_data")) {
  dir.create("bulk_station_data")
}

# Option 1: Sequential processing (original method)
pwalk(
  list(
    station_id = stations_meta$station_id,
    station_name = stations_meta$station_name,
    city_name = stations_meta$city_name
  ),
  function(station_id, station_name, city_name) {
    message("Downloading data for station: ", station_name, " (", station_id, ")")

    # Call download_all_data_for_station() with only the essential inputs.
    # This function automatically uses a broad date range (e.g. 1900-01-01 to 2099-12-31)
    # to fetch all available data for that station.
    station_data <- download_all_data_for_station(
      station_id = station_id,
      parameters = parameters,
      frequency = frequency
    )

    # Optionally, if station_data might be huge, you can periodically flush memory or
    # simply write the result directly.
    file_name <- file.path("bulk_station_data", paste0("data_", station_id, ".parquet"))
    arrow::write_parquet(station_data, sink = file_name, compression = "snappy")

    message("Saved data to: ", file_name)
  }
)

# Option 2: Parallel processing with furrr (much faster!)
plan(multisession, workers = 4)

future_pwalk(
  list(
    station_id = stations_meta$station_id,
    station_name = stations_meta$station_name,
    city_name = stations_meta$city_name
  ),
  function(station_id, station_name, city_name) {
    # Download all available data with internal parallel processing for chunks
    station_data <- download_all_data_for_station(
      station_id = station_id,
      parameters = parameters,
      frequency = frequency,
      parallel = TRUE, # Also parallelize time chunks
      show_progress = FALSE # Disable messages in parallel
    )

    # Save to Parquet file
    file_name <- file.path("bulk_station_data", paste0("data_", station_id, ".parquet"))
    arrow::write_parquet(station_data, sink = file_name, compression = "snappy")

    return(file_name)
  },
  .options = furrr_options(seed = TRUE)
)

plan(sequential) # Reset to sequential processing
```

## Troubleshooting

### Common Issues and Solutions

#### "Invalid station_id format"
**Problem:** Station ID is not in valid UUID format.

**Solution:**
```{r, eval=FALSE}
# Get valid station IDs from list_stations()
stations <- list_stations()
valid_station_id <- stations$station_id[1]

# Use this ID in your downloads
data <- download_air_quality_data(
  station_id = valid_station_id,
  parameters = c("PM10"),
  start_datetime = "2025-03-16 00:00",
  end_datetime = "2025-03-17 00:00"
)
```

#### "start_datetime must be before end_datetime"
**Problem:** Date range is reversed or invalid.

**Solution:**
```{r, eval=FALSE}
# Correct format: "YYYY-MM-DD HH:MM"
start_datetime <- "2025-03-01 00:00" # Earlier date
end_datetime <- "2025-03-31 23:59" # Later date

# Check your dates before downloading
lubridate::ymd_hm(start_datetime) < lubridate::ymd_hm(end_datetime)
```

#### Empty or No Data Returned
**Problem:** Query returns empty tibble.

**Possible causes:**

1. Station has no data for the specified date range
2. Parameters not measured at this station
3. Network connectivity issues
4. API temporarily unavailable

**Solution:**
```{r, eval=FALSE}
# Check if station exists and is active
stations <- list_stations()
station_info <- stations |> filter(station_id == your_station_id)

# Try a recent date range (last 7 days)
data <- download_air_quality_data(
  station_id = your_station_id,
  parameters = c("PM10"),
  start_datetime = format(Sys.time() - 7 * 24 * 60 * 60, "%Y-%m-%d %H:%M"),
  end_datetime = format(Sys.time(), "%Y-%m-%d %H:%M"),
  frequency = "hourly"
)

# Check multiple stations to verify API connectivity
```

#### Slow Downloads
**Problem:** Downloads taking too long.

**Solutions:**
```{r, eval=FALSE}
# 1. Use parallel processing
bulk_data <- bulk_download_air_quality_data(
  start_datetime = "2025-03-16 00:00",
  end_datetime = "2025-03-16 12:00",
  parameters = c("PM10"),
  parallel = TRUE,
  workers = 4
)

# 2. Reduce date range
# Instead of years, download months at a time

# 3. Use daily instead of hourly frequency
data <- download_air_quality_data(
  station_id = station_id,
  parameters = c("PM10"),
  start_datetime = "2024-01-01 00:00",
  end_datetime = "2024-12-31 23:59",
  frequency = "daily" # Much smaller dataset
)

# 4. Download fewer parameters at once
```

#### "Package 'future' not found" (Parallel Processing)
**Problem:** Trying to use parallel processing without required packages.

**Solution:**
```{r, eval=FALSE}
install.packages(c("future", "furrr", "progressr"))
library(future)
library(furrr)
```

#### Network Timeouts or Connection Errors
**Problem:** API requests timing out.

**Solutions:**

1. Check internet connection
2. Try again later (API might be under maintenance)
3. Reduce number of parallel workers
4. Use smaller date ranges or fewer parameters

```{r, eval=FALSE}
# Reduce parallel workers if getting timeouts
data <- bulk_download_air_quality_data(
  start_datetime = "2025-03-16 00:00",
  end_datetime = "2025-03-16 12:00",
  parameters = c("PM10"),
  parallel = TRUE,
  workers = 2 # Reduced from 8
)
```

#### Memory Errors in Parallel Mode
**Problem:** Out of memory when using parallel processing.

**Solutions:**
```{r, eval=FALSE}
# 1. Reduce number of workers
workers <- 2 # Instead of 8

# 2. Process in smaller batches
# 3. Write results to disk immediately (use Parquet/Arrow)
# 4. Use daily frequency instead of hourly
```

#### Progress Bar Not Showing
**Problem:** No progress indication during downloads.

**Solution:**
```{r, eval=FALSE}
# For sequential downloads, ensure cli is installed
install.packages("cli")

# For parallel downloads, use progressr
install.packages("progressr")
library(progressr)

handlers(global = TRUE)
handlers("progress")

with_progress({
  data <- bulk_download_air_quality_data(
    start_datetime = "2025-03-16 00:00",
    end_datetime = "2025-03-16 12:00",
    parameters = c("PM10"),
    parallel = TRUE,
    workers = 4,
    show_progress = TRUE
  )
})
```

### Getting More Help

If you're still experiencing issues:

1. **Check function documentation**:
   ```{r, eval=FALSE}
?download_air_quality_data
?bulk_download_air_quality_data
?list_stations
   ```

2. **Review vignettes** for detailed examples:
   ```{r, eval=FALSE}
vignette("getting-started", package = "airqualitytr")
   ```

3. **Report bugs**: https://github.com/emraher/airqualitytr/issues
   - Include your R version: `R.version.string`
   - Include package version: `packageVersion("airqualitytr")`
   - Provide a minimal reproducible example
   - Include any error messages

## Data Source and Citation

### Data Source
This package accesses air quality monitoring data from the **Turkish Ministry of Environment, Urbanization and Climate Change** (Çevre, Şehircilik ve İklim Değişikliği Bakanlığı).

**Official Data Portal:** <https://sim.csb.gov.tr/>

### Licensing and Usage
**Important:** Before using this data in research or publications, please:

1. **Check the official data policy** at the Ministry's website for:
   - Data licensing terms
   - Citation requirements
   - Usage restrictions and permissions
   - Redistribution rights

2. **Cite appropriately** in your work:
   - Cite the data source (Turkish Ministry of Environment)
   - Cite this R package if used for data access
   - Follow any specific citation requirements from the Ministry

### Suggested Citation Format

**For the data:**
```
Turkish Ministry of Environment, Urbanization and Climate Change. (Year).
Air Quality Monitoring Data. Retrieved from <https://sim.csb.gov.tr/>
```

**For this package:**
```{r, eval=FALSE}
citation("airqualitytr")
```

### Data Quality and Limitations

Please note:

- Data is provided as-is from the Ministry's monitoring network
- This package does not validate or quality-control the data
- Users are responsible for their own data quality checks
- Station metadata and availability may change over time
- Historical data availability varies by station
- Some stations may have gaps or missing measurements

**Always perform your own data quality assessment before analysis.**

## Code of Conduct
  
  Please note that the airqualitytr project is released with a [Contributor Code of Conduct](https://eremrah.com/airqualitytr/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
